An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-05-06 cal_demo/ms.tex
more progress towards addressing reviewer comments

\newcommand{\Nant}{\ensuremath{N_{\mathrm{a}}}}
\newcommand{\Ng}{\ensuremath{N_{\mathrm{g}}}}
\newcommand{\s}{\ensuremath{\hat{\mathbf{s}}}} % s-hat for sine-projected direction
\newcommand{\spix}{\ensuremath{\hat{\mathbf{s}}_{0}}}
\newcommand{\Cna}[1][n]{\ensuremath{\mathcal{C}^{(#1)}_{a,\spix}}}
\newcommand{\ri}{\ensuremath{\mathbf{r}_i}}
\newcommand{\ra}{\ensuremath{\mathbf{r}_a}}
\newcommand{\rb}{\ensuremath{\mathbf{r}_b}}
\newcommand{\beamr}{\ensuremath{\widetilde{W}}}
\newcommand{\beamtheta}{\ensuremath{W}}
\newcommand{\Er}[1]{\ensuremath{\widetilde{E}_{#1}}}
\newcommand{\Erest}[1]{\ensuremath{\widetilde{E}'_{#1}}}
\newcommand{\Ethetaest}{\ensuremath{E'}}
\newcommand{\V}{\ensuremath{\widetilde{V}}}
\newcommand{\dif}{\mathrm{d}}
\newcommand{\caliter}{400}
\newcommand{\damp}{\ensuremath{\gamma}}
\newcommand{\itr}{20}
\newcommand{\tcal}{\ensuremath{t_{\mathrm{cal}}}}
\newcommand{\teff}{\ensuremath{t_{\mathrm{eff}}}}

@@

As a starting point, we consider the feedback calibration outlined in \citealt{mor11}. There it 
was suggested to form a correlation of the uncalibrated antenna measurements with the complex conjugate of an image 
pixel from the output of the correlator. This is a statistically stationary quantity, which can be 
related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum 
needed to calibrate a simple sky of a single point source. However, we aim for a more 
generalized solution for arbitrarily complex sky models. We therefore study the full expression 
for the antenna-pixel correlation,
\begin{equation}\label{eq:Cna}
\Cna \equiv \left<E_a^{(n)}(t) \mathcal{E}'^{*(n)}(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ 
calibration loop, 
the superscript $^*$ indicates complex conjugation,
and \spix\, is the pixel center nearest a bright calibrator of interest. The 
following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which 
contains a bright source to achieve a high signal to noise.

Plugging equation \ref{eq:epix} into equation \ref{eq:Cna}, 
and substituting the $g^{(n)}_b$ for $g'_b$ to indicate the specific loop number,
we find,
\begin{align}\label{eq:cna}
\Cna & = \left<\left(g_a E_a^I(t)+n_a(t)\right) \frac{1}{\Nant} \right. \nonumber \\
& \qquad \left. \times \sum_b \frac{1}{g^{*(n)}_b}\left(g^*_bE_b^{I*}(t)+n^*_b(t)\right) e^{-2\pi i \spix \cdot \rb} \mathcal{W}^*_b(\spix)\right>_t \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \nonumber \\
& \qquad \times \left<\left(g_aE_a^I(t)+n_a(t)\right)\left(g^*_bE_b^{*I}+n^*_b(t)\right) \right>_t \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \nonumber \\
& \qquad \times \left(g_a g^*_b \left<E^I_a(t)E^{*I}_b(t)\right>_t + \left<n_a(t) n^*_b(t)\right>_t\right) \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \left(g_ag^*_bV^I_{ab}+n_{ab}\right)
\end{align}
where in the second step we group time-dependent terms, 
in the third step we assume the noise does not correlate with the ideal electric field 
measurements, and in the final step we introduce the baseline noise $n_{ab}\equiv
\left<n_a(t)n^*_b(t)\right>_t$, and ideal visibilities as the correlation between ideal
electric field measurements, $V^I_{ab} \equiv\left<E^I_a(t)E^{*I}_b(t)\right>_t$.
To be clear, visibilities are never directly formed from the data. Rather it is convenient to
express $\Cna$ in terms of the ideal visibilities, which we can model when solving for the
gains below.

\textcolor{red}{need a brief comment on the noise term}

The next step is to use this antenna-pixel correlation to update our gain solution. We will model 
the right hand side of equation~\ref{eq:cna} by assuming our current gain estimates are 
approximately correct, $g^{(n)}_b\approx g^T_b$, and replacing the true visibilities with a set of 
model visibilities, $\V'_{ab}$. These model visibilities rely an knowledge of the sky, and should 
be precomputed offline. We can then solve equation~\ref{eq:cna} for $g^T_a$ to achieve an 
updated estimate.
\begin{equation}\label{eq:cal_solution}
g'^{(n+1)}_a = \Cna \Nant \left[ \sum_b \beamtheta^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \V'_{ab} \right]^{-1}
\end{equation}
This equation is our prescription for estimating the antenna gains of a direct imaging array. The 
online computation complexity scales as $\mathcal{O}(\Nant)$ as we form a $\Cna$ for each 
antenna.

>>>>

\newcommand{\Nant}{\ensuremath{N_{\mathrm{a}}}}
\newcommand{\Ng}{\ensuremath{N_{\mathrm{g}}}}
\newcommand{\s}{\ensuremath{\hat{\mathbf{s}}}} % s-hat for sine-projected direction
\newcommand{\spix}{\ensuremath{\hat{\mathbf{s}}_{0}}}
\newcommand{\Cna}[1][n]{\ensuremath{\mathcal{C}^{(#1)}_{a,\spix}}}
\newcommand{\Kna}[1][n]{\ensuremath{\mathcal{K}^{(#1)}_{a,\spix}}}
\newcommand{\ri}{\ensuremath{\mathbf{r}_i}}
\newcommand{\ra}{\ensuremath{\mathbf{r}_a}}
\newcommand{\rb}{\ensuremath{\mathbf{r}_b}}
\newcommand{\beamr}{\ensuremath{\widetilde{W}}}
\newcommand{\beamtheta}{\ensuremath{W}}
\newcommand{\Er}[1]{\ensuremath{\widetilde{E}_{#1}}}
\newcommand{\Erest}[1]{\ensuremath{\widetilde{E}'_{#1}}}
\newcommand{\Ethetaest}{\ensuremath{E'}}
\newcommand{\V}{\ensuremath{\widetilde{V}}}
\newcommand{\dif}{\mathrm{d}}
\newcommand{\caliter}{400}
\newcommand{\damp}{\ensuremath{\gamma}}
\newcommand{\itr}{20}
\newcommand{\tcal}{\ensuremath{t_{\mathrm{cal}}}}
\newcommand{\teff}{\ensuremath{t_{\mathrm{eff}}}}

@@

As a starting point, we consider the feedback calibration outlined in \citealt{mor11}. There it 
was suggested to form a correlation of the uncalibrated antenna measurements with the complex conjugate of an image 
pixel from the output of the correlator. This is a statistically stationary quantity, which can be 
related to the sum of visibilities involving the antenna used in the correlation -- exactly the sum 
needed to calibrate a simple sky of a single point source. However, we aim for a more 
generalized solution for arbitrarily complex sky models. We therefore study the full expression 
for the antenna-pixel correlation,
\begin{equation}\label{eq:Cna}
\Kna \equiv \left<E_a^{(n)}(t) \mathcal{E}'^{*(n)}(\spix,t)\right>_t,
\end{equation}
where the superscript $n$ again represents the quantity formed in the $n^\mathrm{th}$ 
calibration loop, 
the superscript $^*$ indicates complex conjugation,
and \spix\, is the pixel center nearest a bright calibrator of interest. The 
following will hold for any chosen pixel, \spix, though it is advantageous to choose a pixel which 
contains a bright source to achieve a high signal to noise.

Plugging equation \ref{eq:epix} into equation \ref{eq:Cna}, 
and substituting the $g^{(n)}_b$ for $g'_b$ to indicate the specific loop number,
we find,
\begin{align}
\Kna & = \left<\left(g_a E_a^I(t)+n_a(t)\right) \frac{1}{\Nant} \right. \nonumber \\
& \qquad \left. \times \sum_b \frac{1}{g^{*(n)}_b}\left(g^*_bE_b^{I*}(t)+n^*_b(t)\right) e^{-2\pi i \spix \cdot \rb} \mathcal{W}^*_b(\spix)\right>_t \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \nonumber \\
& \qquad \times \left<\left(g_aE_a^I(t)+n_a(t)\right)\left(g^*_bE_b^{I*}+n^*_b(t)\right) \right>_t \nonumber \\
& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \nonumber \\
& \qquad \times \left(g_a g^*_b \left<E^I_a(t)E^{*I}_b(t)\right>_t + g_a\left<E_a^I(t)n^*_b(t)\right>_t \right. \nonumber \\
& \qquad \left. + g^*_b\left<n_a(t)E^{I*}_b(t)\right>_t + \left<n_a(t) n^*_b(t)\right>_t\right)  
%& = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} \left(g_ag^*_bV^I_{ab}+n_{ab}\right)
\end{align}
where in the second step we group time-dependent terms, 
and in the third step we expanded the product within the time average.
Because the noise does not correlate with the ideal antenna measurements we can
drop the middle terms. We will also assume that the noise of antenna $a$ correlated with
the noise of antenna $b$ is also zero mean, unless $a=b$. By defining an ideal visibility as
$V_{ab}^I \equiv \left<E^I_a(t)E^{*I}_b(t)\right>_t$, we can simplify to
\begin{align}\label{eq:Kna}
\Kna & = \frac{1}{\Nant} \sum_b \frac{1}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb}\nonumber \\
& \qquad \times \left( g_ag_b^* V_{ab}^I + \delta_{ab}\left<|n_a(t)|^2\right>_t\right)
\end{align}
where $\delta_{ab}$ is the Kronecker-delta function. To be clear, visibilities are never directly 
formed from the data. Rather it is convenient to express $\Kna$ in terms of the ideal 
visibilities, which we can model when solving for the gains below.

We will ultimately wish to model the right side of this equation to in order to solve for the
antenna gains. However, the self-correlated noise bias, $\left<|n_a(t)|^2\right>$, can often be 
significantly larger than the cross correlation power, and difficult to model precisely. It 
is therefore beneficial to subtract the self-correlation directly from $\Kna$. By properly
accounting for the various weighting and phasing terms, we can use self-correlation of the
uncalibrated antenna electric field measurements to remove the noise bias and $V^I_{aa}$
term from equation~\ref{eq:Kna}.
\begin{align}\label{eq:cna}
\Cna & \equiv \Kna - \frac{1}{\Nant g^{*(n)}_a}\mathcal{W}^*_a(\spix)e^{-2\pi i \spix \cdot \ra} \left<|E_a|^2\right>_t\nonumber \\
& = \frac{g_a}{\Nant} \sum_{b\ne a} \frac{g^*_b}{g^{*(n)}_b} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} V^I_{ab}
\end{align}

The next step is to use this antenna-pixel correlation to update our gain solution. We will model 
the right hand side of equation~\ref{eq:cna} by assuming our current gain estimates are 
approximately correct, $g^{(n)}_b\approx g_b$, and replacing the ideal visibilities with a set of 
model visibilities, $V^M_{ab}$. These model visibilities rely on knowledge of the sky, and should 
be precomputed before observing. We can then solve equation~\ref{eq:cna} for $g_a$ to achieve an 
updated estimate.
\begin{equation}\label{eq:cal_solution}
g^{(n+1)}_a = \Cna \Nant \left[ \sum_{b\ne a} \mathcal{W}^*_b(\spix) e^{-2\pi i \spix \cdot \rb} V^M_{ab} \right]^{-1}
\end{equation}
This equation is our prescription for estimating the antenna gains of a direct imaging array. The 
real-time computation complexity scales as $\mathcal{O}(\Nant)$ as we form 
an antenna-pixel correlation, $\Cna$, for each 
antenna.

\textcolor{red}{pick up here}
